JD notes on effects

### 2022-07-05 manuscript/body.tex
After JD, BB and BC meeting. Going for tea

Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model. These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change. In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.

To make an outcome plot, we use a \emph{focal} predictor on the x-axis. The resulting plot will also depend on choices we make about other (non-focal) predictor variables. In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate. However, when the focal predictor has interactions, or when there is a non-linear link function, non-focal choices can also affect the slope of the response. Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms) which extend ordinary linear models to incorporate random effects. How we treat random effects can also the estimated response.

\jd{We should discuss before continuing this part.}
\BC{Okay. Some thoughts below your \P}
The outcome plots described above are often loosely called either prediction plots or effects plots. We endeavor here to make a conceptual distinction. If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central (outcome) estimate which is the same for both prediction and effects plots. \jd{This \P\ should be about distinguishing -- what do they mean? I feel we can make good progress after a terminology meeting.}. It simply tells us what we expect the outcome to be for any given value of the focal input variable. We may also want to provide uncertainty associated with the central estimate. How we choose to describe the uncertainty can inform the choice we make between prediction and effects plots. For instance, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects) or we may be interested in showing the uncertainty associated with the focal predictor only. In the former case, we use the prediction plots which generally focus on the \emph{total} effect, and so will often be suited for a univariate model. On the other hand, in the latter case, we introduce effects plots which focus on the direct effect of the focal predictor -- after controlling for other non-focal input variables -- from a multivariate model, and so will be preferable for multivariate models. Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots.} \BC{Not sure whether to say that uncertainties are described by the CIs?}

Generating quantities, i.e., central estimate together with the corresponding confidence intervals, associated  prediction and effects plots has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal input variable and the \emph{reference point} for non-focal input variables, especially in multivariate models
\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms).
\end{enumerate}

\jd{Not sure I agree with this sentence. Also, picking representative focal values seems not interesting? The reason I'm not sure I agree is that our method of spanning the range with a fixed number of points is not exactly a quantile method; also, what does "sized" mean here?}
\BC{I somehow agree it is not very interesting but I think we need to mention what we do. We do both -- spanning a range of values and quantiles of the observed focal values (our default). "sized" is supposed to the number of sampled data points. Maybe I should just say we take quantiles of the focal predictor?.}

The most common way of choosing the representative values is taking unique levels or quantiles \jd{I still don't understand where quantiles come in? What's an example of the use of quantiles?} of the focal input variable if discrete or continuous, respectively.  The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. 
\BC{We need to decide on how to proceed from here. I think we can refer to this as the central estimate. Re-writing and we can delete older version if that's okay.}
\jd{I had a suggestion in the notes.}
In this article, we refer to these quantities as the central (outcome) estimates for an ``average case'' in a population; and the associated confidence intervals will either be \jd{WHAT?? a prerequisite for} prediction or effects plots. In the case of effects plots, we can also choose an anchor (value chosen for the focal input variable). The concept of effects plots has been presented in the literature -- Shi et al. \citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.
\jd{This is useful text, but also feels repetitive; why are we coming back to the “concept” of effects plots?}

\BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}

For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging. As a result the bias in the expected (mean) predictions can easily be corrected by performing predictions from the \emph{model center} (average of the predictor variables). The mean prediction is the prediction at the model center. On the other hand, when dealing with GL(M)Ms with nonlinear link functions, the averaging is done on the nonlinear scale, i.e., nonlinear averaging, making estimating correct predictions much harder. This is the \emph{bias} in expected mean prediction induced by nonlinear transformation of the response variable and it lead to inaccurate predictions. One way to deal with this is to make predictions on the transformed scale (linear predictor scale), and then back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. In linear models, mean-based gives estimates consistent with the observed values, however, in models with nonlinear link functions, this is not always the case. An alternative to mean-based reference point, i.e., averaging non-focal predictor variables, is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictor variables and then averaging across the values of the focal predictor variable \citep{hanmer2013behind}.

The main purpose of this article is to discuss and implement various approaches for computing predictions and provide an alternative method for computing the associated confidence intervals, i.e., effects. We further explore and demonstrate, using simulated data, approaches for correcting bias in predictions for GL(M)Ms involving nonlinear link functions. The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.

\jd{This seems like too much about the other packages. Maybe we could: 1) stake things out a bit better by closing off our overview with the "splash plot" (make that the first figure); 2) write the other-package stuff specifically as a comparison, and make it a bit shorter.}
\BC{By ``splash plot'' you mean Fig 1? I am suggesting we move all the package comparison stuff to/after statistical background section or maybe to simulation example section ...}

@@

\item \textbf{Focal and non-focal input variables}: The focal input variable refers to the input variable we are interested in. For multivariate models, all other variables other than focal input variable are non-focal input variables. For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal input variable while $x_2$ and $x_3$ are non-focal input variables.

@@

\item \textbf{Predictor variables (or simply predictors)}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix. Each input variable may correspond to one or more predictor variables. In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one predictor variables. For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ predictors -- $x_1, x_2, x_3$ plus intercept term and $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
\item \textbf{Focal predictors:} All columns of the model matrix associated with the focal input variable. For example, all the terms of a cubic polynomial focal variable or columns corresponding to a categorical focal variable with more than two levels.
\item \textbf{Non-focal predictors:} All columns of the model matrix associated with the non-focal input variable(s). For example, in the previous household size example, all the columns associated with religion are non-focal predictors if income is the focal input variable.
\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more predictor variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of predictor variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point. If there are more than one predictor variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.

@@

In whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging. In other words, the averaging is no longer on the link scale and are likely to be bounded by the original data scale.

\jd{What does this mean? I was going to say that $g$ collapses out of \EREF{compare_anchored_pop_based} and the results are identical. Is that wrong?} As stated above, in simple linear models without interactions, the effect is constant, so both approaches yield similar results. 
However, picking a single value, e.g., mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic or not contained in or representative of the population. In addition, the mean-based approach fails to use every values of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach, summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This is might not be a problem in whole-sample-based approach since it focuses on specific observations because the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.

\jd{I meant to say earlier that we need to talk about the common “sum-to-zero-like" method of averaging over dsitnct values of a categorical input variable vs.~the true model-center method. This might be another place where we can show we do better than some existing packages! See notes.}
Another potential concern with the mean-based approach arises in situations where direct naive use leads to rare or meaningless basis for generalization. For example, setting dummy categorical variables in the model matrix to their means, which, by default, sets them to their sample means or observed proportions. For example, a dummy variable for christian household heads may be set to $0.2$, translating to prediction for an household head who is $20\%$ christian \citep{hanmer2013behind}.

Whole-sample-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, whole-sample-based approach can be computationally intensive for large datasets.

>>>>

Plots of predicted values of an outcome against predictors (often called effect plots or prediction plots) are often a useful way to summarize the results of a regression model. These can be used to illustrate model uncertainty or to give a more explicit quantitative sense of how the outcome is expected to change. In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in understanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.

To make an outcome plot, we use a \emph{focal} predictor on the x-axis. The resulting plot will depend on choices we make about other (non-focal) predictor(s). In ordinary (Gaussian) linear regression, the non-focal choices may have a simple additive effect on the outcome estimate. However, when the focal predictor has interactions or a non-linear link function, non-focal choices can also affect the slope of the response. Additional challenges arise when dealing with generalized linear (mixed) models (GL(M)Ms), which extend ordinary linear models to incorporate random effects. How we treat random effects can also have an effect.

The outcome plots described above are often called prediction plots or effects plots. We endeavor here to make a conceptual distinction. If our goal is to \emph{predict} what we learn about the outcome variable by measuring the focal predictor, \NEW{then we would focus on the central estimate, which is the same for both prediction and effects plots. However, the distinction between the two lies in how we describe the uncertainties around the central estimate. In particular, we may want to capture the uncertainty in the whole model (these include uncertainty due to intercept, focal and non-focal predictors, and random effects), or we may be interested in showing the uncertainty associated with the focal predictor only. In the former case, we use the prediction plots, which generally focus on the \emph{total} effect and will often be suited for a univariate model. In the latter case, we use effect plots that focus on the direct effect of the focal predictor -- after controlling for other non-focal predictors -- from a multivariate model, which will be preferable for multivariate models. Since effects plots focus on focal-specific uncertainty, they are usually narrower than prediction plots. Shi et al. \citep{shi_evidence_2017} used effects plots to compare the difference in sexual risk behaviors between circumcised and uncircumcised men.}

Generating quantities, i.e., central estimate, together with the corresponding confidence intervals, i.e.,  prediction and effects plots, has some challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
\item uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
\item bias in the expected mean prediction induced by the non-linear transformation of the response variable (especially in GL(M)Ms).
\end{enumerate}

The most common way of choosing the representative values is taking unique levels or quantiles of the focal predictor if discrete or continuous, respectively. The central estimates are then calculated by holding the non-focal predictors at their reference point (values chosen for the non-focal predictors, e.g., means) while varying the focal predictor, with the goal that the estimates represent how the model responds to the changes in the focal predictor \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as the \emph{central estimates} for an ``average case'' in a population; the associated confidence intervals will either describe a prediction or effect plot. In the effect plots, we can also choose an anchor (value chosen for the focal predictor).

For linear models (including models with complex interactions), the averaging is done on the linear scale, i.e., linear averaging. As a result, the \emph{bias} in the expected (mean) predictions can easily be corrected using \emph{mean-based reference point}, i.e., averaging of the non-focal model variables -- currently not implemented in commonly used \proglang{R} software packages. The mean of the central estimates is the prediction at the \emph{model center}. In this case, the estimates are usually consistent with the observed values. However, in models with non-linear link functions, this is not always the case. In particular, when dealing with GL(M)Ms with non-linear link functions, the averaging is done on the non-linear scale, i.e., non-linear averaging, making generating correct predictions much harder. This is the bias in the expected mean prediction induced by the non-linear transformation of the response variable. One way to address this is to make predictions on the linear predictor scale and back-transform to the original scale. However, the back-transformation may either result in biased predictions or requires some approximation, for example, second-order Taylor approximation implemented in \pkg{emmeans} \citep{lenth2018package}. An alternative to the mean-based reference point is the \emph{whole-sample-based} approach, discussed in detail later, which involves computing the prediction over the population of non-focal predictors and then averaging across the values of the focal predictor \citep{hanmer2013behind}.

This article aims to discuss and implement various approaches for computing predictions and effects, together with the associated plots. We further explore and demonstrate, using simulated data, approaches for correcting bias in central estimates for GL(M)Ms involving non-linear link functions. The proposed method and \proglang{R} software package will complement the existing ones by providing: 1) a straightforward way to generate effects, and 2) an alternative and a more robust way to correct for prediction bias in GL(M)Ms.

@@

\item \textbf{Focal and non-focal predictors}: The focal predictor refers to the input variable we are interested in. For multivariate models, all other variables other than focal predictors are non-focal predictors. For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.

@@

\item \textbf{Model variables}: Refer to the X-variables which go into the regression model; and represent columns in the model matrix. Each input variable may correspond to one or more model variable. In particular, variables with more than two categories (for instance religion in our previous example), or input variables modeled with interactions, a spline or polynomial interactions, will correspond to more than one model variable. For example, the regression models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher} have $4$ model variables -- $x_1, x_2, x_3$ plus intercept term and $5$ model variables -- $x_1, x_2, x_3, x_2x_3$ plus the intercept term, respectively.
\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach that uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}B. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for the center point. If there are more than one model variables associated with the focal input variable, however, the model center does not correspond to any given value of the variable, and we do not necessarily get the ``nice'' crossing we get in simpler cases, see Fig~\ref{fig:pred_cubic_plots}.

@@

In the whole-sample-based approach, the ensemble of predictions and CIs are back-transformed before averaging, see Equation~\ref{eq:pop_response}, so we do not need to worry about the nonlinear averaging. In other words, the averaging is no longer on the link scale and is likely to be bounded by the original data scale. In simple linear models without interactions, averaging on the link scale is identical to averaging on the response, so both approaches yield similar results. However, picking a single value, e.g., the mean of the predictor, on which to draw conclusions about the effect can be problematic, unrealistic, or not contained in or representative of the population. In addition, the mean-based approach fails to use every value of non-focal predictors hence not utilizing the full potential of the information contained in the data. This may limit the inferences we can make about the entire population. In general, the mean-based approach provides the predictions of an average case, whereas the whole-sample-based approach summarizes the predictions over the entire population. In some applications, the effect of an average case might not be generalizable to the entire population, especially, if the average does not represent the population. This might not be a problem in the whole-sample-based approach since it focuses on specific observations -- the prediction is first obtained for each observation and then averaged across the levels of the focal input variable.

Another potential concern with the mean-based approach arises when direct naive use leads to a rare or meaningless basis for generalization. For example, if our sample has 20\% Jews, 30\% Muslims, and 50\% Christians. One approach (default in common packages) is assigning equal category weight, i.e., 1/3 Jews, 1/3 Muslims, and 1/3 Christians (the ``sum-to-zero'' approach). The second approach (our default) is setting dummy categorical variables in the model matrix to their means, which, by default, set them to their sample means or observed proportions. The first or second approach translates to prediction for a household head who is 1/3 or 50\% Christian, respectively \citep{hanmer2013behind}. The second approach seems more realistic and will converge to the population mean in many cases.

The whole-sample-based approach is not entirely foolproof. For instance, similar to the mean-based approach, in the case of continuous focal predictors, choosing the representative values of the focal predictors can be very challenging, especially if the cases are not evenly distributed around the minimum and the maximum values or within some subgroups defined in the population. In addition, the whole-sample-based approach can be computationally intensive for large datasets.
